---
title: "Maximum Likelihood Fits to the Malaria Therapy Data"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r, echo=FALSE, purl=FALSE}
library(knitr)
library(deSolve)
library(viridisLite)
library(viridis)
```

```{r}
library(ramp.falciparum)
```

In the following, we fit a model to the parasite counts from the malaria-therapy data. This document walks through the MLE algorithm. 

## Malaria Therapy

We read in the malaria therapy data:  

```{r read data}
malariatherapy = read.csv("MT_PT_NP.csv")
```

The function `getOne` pulls all the data from a single patient and returns an object with two columns: 

1. `alpha` -- The age of an infection, in days 

2. `B` -- The parasite count 

```{r getOne}
getOne = function(i, dta){
   ix = which(!is.na(dta[,i]))
   ix = ix[1:max(which(dta[ix,i]>0))]
   cbind(alpha=ix+20, B=dta[ix,i])
}
vals = getOne(1, malariatherapy)
head(vals) 
```
We now extract and concatenate all the data into two vectors of paired data: 

1. `alpha_list` is the AoI for all observations 

2. `counts_list` is the observed parasite counts for all observations 

```{r all observations}
for(i in 2:334) vals = rbind(vals,getOne(i, malariatherapy))
alpha_list = vals[,1]
counts_list = vals[,2]
```

## Maximum Likelihood Estimation 

In the following, we present code to fit the parameters for two functions and the observational process: 

1. $F_\mu,$ which describes mean $\log_{10}$ parasite density distributions as a function of the AoI, $\alpha.$ We fit two shape parameters.

2. The parasite density distribution function, $\Omega(\mu).$ We fit parameters that describe the variance as a function of the mean. 

3. We assume parasite counts follow a negative binomial distribution. 

### Expected Densities by AoI 

We let $F_\mu$ be a function that describes expected $\log_{10}$ densities, given the age of infection $\alpha$ 

$$
F_\mu(\alpha) = 
\begin{cases}
    \text{NA} \ & \mbox{if } 0 \leq \alpha < 7 \\[10pt]
      l + (b - l)  \frac{{\textstyle \alpha}}{{\textstyle \delta}}   & \mbox{if } 7 \leq \alpha \leq \delta  \\[10pt]
     l + (b - l) e^{-s_\alpha (\alpha-\delta)}   & \mbox{if } \alpha \geq \delta\\  
\end{cases}
$$
In `ramp.falciparum,`  this function is called `Fmu.base.` A function to configure the function is `par_Fmu_base.`

```{r}
aa = seq(7, 250)
par1 <- par_Fmu_base()
plot(aa, Fmu(aa,0, par1), type = "l", xlab = expression(alpha), ylab = expression(F[mu]))
```

We want to fit values of $s_\alpha$ and $b$

During fitting, random values could lead to nonsensical values, so we write a function $F_b$ to bound the value of $b$: 

```{r F_b}
F_b = function(x, bvm){
  1.5 + exp(.2*x)/(1+exp(.2*x))*(bvm-3)
}
```

Values between 1.5 and 11.5 are allowed: 

```{r}
c(F_b(-100, 13),F_b(100, 13))
```

Similarly, we want to constrain the parameter $s_\alpha$ to be between 0 and 1: 

```{r F_Sa} 
F_Sa = function(x){
  exp(x)/(1+exp(x))
}
```

```{r}
c(F_Sa(-100),F_Sa(100))
```

### Parasite Density Distribution Function 

We let $\Omega(\mu)$ denote the density of parasites, given an expected value $\mu.$

In the following, we let $\Omega(\mu)$ be a modified $\beta$ distribution. The variance must be less than $\mu (1-\mu).$  We use the function 

$$ 
\sigma(\mu) = \mu^{1+b_\sigma}(1-\mu)^{1+a_\sigma}
$$

```{r}
pO <- par_sigma_abc(aa = 3.11, bb=2.14, cc=1)
pB <- par_Omega_beta(pO)
```

```{r}
clrs = turbo(5)
xi = seq(1, 12, length.out=200)
plot(xi, d_Omega(xi, 10, par_Omega=pB), type = "l", col = clrs[1], 
     xlab = expression(xi), 
     ylab = expression(Omega) )
lines(xi, d_Omega(xi, 9, par_Omega=pB), col =clrs[2])
lines(xi, d_Omega(xi, 8, par_Omega=pB), col =clrs[3])
lines(xi, d_Omega(xi, 7, par_Omega=pB), col =clrs[4])
lines(xi, d_Omega(xi, 6, par_Omega=pB), col =clrs[5])
```


### Counts 

Here, we assume the counts follow a negative binomial distribution, parameterized with **mu** and **size**. 

## Sampling 

The sampling process happens in two steps: 

1. First, we compute the expected value for the count given the age of the infection.

2. Next, we specify a probability distribution function and compute the likelihood of the observed count.

### Expected Value for the Count

We write a function to compute the log-likelihoood of a single observation, a pair of observations $\alpha_i$ and a count $\hat \xi_i.$ The expected value for the count is a convolution over all possible values of $\xi$: 

$$\int_0^b \Omega(\xi | F_\mu(\alpha_i)) C(\hat \xi_i | \xi) d\xi$$

We write a function to generate an initial set of guesses: 

```{r pinit}
pinit = function(){
  c(tildecounts=rnorm(1, 10.25, .2), 
    Sa=rlnorm(1, log(1/300), .1), 
    aa=rnorm(1, 3, .2), 
    bb=rnorm(1, 0.4, .05))
}
pinit() 
```

## Examples

Finally, we can do the fits: 

```{r}
counts_MLE_sz(alpha_list, counts_list, pinit(), 5,   F_b, F_Sa) -> fits1
counts_MLE_sz(alpha_list, counts_list, pinit(), 2,   F_b, F_Sa) -> fits2
counts_MLE_sz(alpha_list, counts_list, pinit(), 1,   F_b, F_Sa) -> fits3
counts_MLE_sz(alpha_list, counts_list, pinit(), 0.5, F_b, F_Sa) -> fits4
counts_MLE_sz(alpha_list, counts_list, pinit(), 0.2, F_b, F_Sa) -> fits5
```

```{r}
fits1[5] <- 5
fits2[5] <- 2
fits3[5] <- 1
fits4[5] <- 0.5
fits5[5] <- 0.2
```

```{r}
fits <- data.frame(rbind(fits1, fits2, fits3, fits4, fits5)) 
```

```{r}
par_F1 = par_Fmu_base(tildeb = fits1[1], Sa=fits1[2])
par_F2 = par_Fmu_base(tildeb = fits2[1], Sa=fits2[2])
par_F3 = par_Fmu_base(tildeb = fits3[1], Sa=fits3[2])
par_F4 = par_Fmu_base(tildeb = fits4[1], Sa=fits4[2])
par_F5 = par_Fmu_base(tildeb = fits5[1], Sa=fits5[2])
```

```{r}
par_O1 = par_Omega_beta(par_sigma_abc(aa=fits1[3], bb=fits1[4]))
par_O2 = par_Omega_beta(par_sigma_abc(aa=fits2[3], bb=fits2[4]))
par_O3 = par_Omega_beta(par_sigma_abc(aa=fits3[3], bb=fits3[4]))
par_O4 = par_Omega_beta(par_sigma_abc(aa=fits4[3], bb=fits4[4]))
par_O5 = par_Omega_beta(par_sigma_abc(aa=fits5[3], bb=fits5[4]))
```

```{r}
source("Figure5.R")
```

```{r}
alpha2PD() -> pdO 
alpha2PD(par_Fmu=par_F1, par_Omega = par_O1) -> pdO1 
alpha2PD(par_Fmu=par_F2, par_Omega = par_O2) -> pdO2 
alpha2PD(par_Fmu=par_F3, par_Omega = par_O3) -> pdO3 
alpha2PD(par_Fmu=par_F4, par_Omega = par_O4) -> pdO4 
alpha2PD(par_Fmu=par_F5, par_Omega = par_O5) -> pdO5 
```

```{r}
surfplot = function(nclrs, pdObj){
  clrs = rev(magma(nclrs))

  with(pdObj, filled.contour(x, y, pd, 
                           xlab = expression(list(alpha, "Parasite Age (in Days)")), 
                           ylab = expression(f[P](xi, alpha)), 
                           xaxt = "n", 
                           yaxt = "n", 
                           nlevels=nclrs, 
                           col = clrs))
}
```

```{r}
surfplot(34, pdO)
```

```{r}
surfplot(34, pdO1)
```
```{r}
surfplot(34, pdO2)
```

```{r}
surfplot(34, pdO3)
```

```{r}
surfplot(34, pdO4)
```
```{r}
surfplot(34, pdO5)
```

```{r Fit it, eval=F}
for(i in 1:10){
  counts_MLE(alpha_list, counts_list) -> fits1
  fits <- rbind(fits, fits1)
} 
```

This was run, originally, with and without simulated annealing, and the results were saved to a file. 

```{r, eval=F}
write.table(fits, "fits.Rtbl")
```

```{r}
fits_from_file <- read.table("fits.Rtbl")
fits_from_file <- data.frame(fits_from_file)
```


