---
title: "Maximum Likelihood Fits to the Malaria Therapy Data"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---



```{r, echo=FALSE, purl=FALSE}
library(knitr)
library(ramp.falciparum)
library(deSolve)
library(viridisLite)
library(viridis)
```

## Malaria Therapy

We read in the malaria therapy data:  

```{r}
library(ramp.falciparum)
```

```{r}
malariatherapy = read.csv("MT_PT_NP.csv")
```

We create two lists:

1. The age of infection

2. The parasite counts

```{r}
getOne = function(i, dta){
   ix = which(!is.na(dta[,i]))
   ix = ix[1:max(which(dta[ix,i]>0))]
   cbind(alpha=ix+20, B=dta[ix,i])
}
vals = getOne(1, malariatherapy)

for(i in 2:334) vals = rbind(vals,getOne(i, malariatherapy))
alpha_list = vals[,1]
counts_list = vals[,2]
```

## MLE 

In the following, we present code to fit the parameters for two functions and the observational process: 

1. $F_\mu,$ which describes mean $\log_{10}$ parasite density distributions as a function of the AoI, $\alpha.$ We fit two shape parameters.

2. The parasite density distribution function, $\Omega(\mu).$ We fit parameters that describe the variance as a function of the mean. 

3. We assume parasite counts follow a negative binomial distribution. 

### $F_\mu$

We let $F_\mu$ be the function

$$
F_\mu(\alpha) = 
\begin{cases}
    \text{NA} \ & \mbox{if } 0 \leq \alpha < 7 \\[10pt]
      l + (b - l)  \frac{{\textstyle \alpha}}{{\textstyle \delta}}   & \mbox{if } 7 \leq \alpha \leq \delta  \\[10pt]
     l + (b - l) e^{-s_\alpha (\alpha-\delta)}   & \mbox{if } \alpha \geq \delta\\  
\end{cases}
$$
We want to fit values of $s_\alpha$ and $b$

During fitting, random values could lead to nonsensical values, so we write a function $F_b$ to bound the value of $b$: 

```{r constrained b, purl=F}
F_b = function(x, bvm){
  1.5 + exp(.2*x)/(1+exp(.2*x))*(bvm-3)
}
```

Values between 1.5 and 11.5 are allowed: 

```{r}
c(F_b(-100, 13),F_b(100, 13))
```

Similarly, we want to constrain the parameter $s_\alpha$ to be between 0 and 1: 

```{r}
F_Sa = function(x){
  exp(x)/(1+exp(x))
}
```


### $\Omega(\mu)$

In the following, we let $\Omega(\mu)$ be a modified $\beta$ distribution. The variance must be less than $\mu (1-\mu).$  We use the function 

$$ 
\sigma(\mu) = \mu^{1+b_\sigma}(1-\mu)^{1+a_\sigma}
$$

```{r}
F_sigma = function(mu, aa=3.11, bb=2.14){
  pmin(mu^(1+abs(bb))*(1-mu)^(1+abs(aa)), mu*(1-mu)) 
}
```

### Counts 

Here, we assume the counts follow a negative binomial distribution, parameterized with **mu** and **size**. 

## Likelihood 

We write a function to compute the log-likelihoood of a single observation, a pair of observations $\alpha_i$ and a count $\hat \xi_i.$ The probability of the count is a convolution over all possible values of $\xi$: 

$$\int_0^b \Omega(\xi | F_\mu(\alpha_i)) C(\hat \xi_i | \xi) d\xi$$
The log-likelihood for a single observation is computed using `llik_count` 

+ `count` is the count 

+ `mu` is the expected value for `xi` 

+ `sz` is the **size** parameter for `dnbinom`

+ `par_O` is a parameter list to dispatch `d_Omega`

+ `q` is the $\log_{10}$ number of RBCc blood sampled 

+ `bvm` is the $\log_{10}$ total number of RBCs 

```{r}
llik_count = function(count, mu, sz, par_O, q=6, bvm=13){
 
  dB = function(xi, count, mu, q, bvm, sz, par_O){
    pr_xi = d_Omega(xi, mu, bvm, par_O)
    pr_count = dnbinom(count, mu=10^(xi-q), size=sz)
    pr_xi*pr_count
  }
  
  lik = integrate(dB, 0, bvm, 
                  count=count, mu=mu,  
                  q=q, bvm=bvm, 
                  sz=sz, par_O=par_O)$value
  -log(lik)
}
```

The log-likelihood for all counts is computed with `llik_counts` 

+ `prms` is the set of values 

+ `alpha` is a list of the ages of infection 

+ `counts` is the corresponding list of the counts 

+ `par_Fm` is a parameter list to dispatch `Fmu`

+ `par_O` is a parameter list to dispatch `d_Omega`

+ `q` is the $\log_{10}$ number of RBCc blood sampled 

+ `bvm` is the $\log_{10}$ total number of RBCs 

```{r}
llik_counts = function(prms, alpha, counts, par_Fmu, par_O, q=6, bvm=13){
  
  llik_count_i = function(i, alpha, counts, sz, par_Fmu, par_O, q=6, bvm=13){
    mu = Fmu(alpha[i], 0, par_Fmu) 
    llik_count(counts[i], mu, sz, par_O, q, bvm)
  }
  
  
  par_Fmu$tildeb = F_b(prms[1], bvm)
  par_Fmu$Sa = F_Sa(prms[2])
  sz=abs(prms[3])
  
  par_O$pSig$aa=abs(prms[4])
  par_O$pSig$bb=abs(prms[5])
  
  lliks = sapply(1:length(alpha), llik_count_i, 
                 alpha=alpha, counts=counts, 
                 sz=sz, par_Fmu=par_Fmu, par_O=par_O, 
                 q=q, bvm=bvm) 
  sum(lliks)
}
```


The maximum likelihood estimation is computed by `counts_MLE` for a paired set set of observations c(`alpha`, `counts`) 

+ `q` is the $\log_{10}$ number of RBCc blood sampled 

+ `bvm` is the $\log_{10}$ total number of RBCs 

```{r}
counts_MLE = function(alpha, counts, q=6, bvm=13){
  par_Fmu = par_Fmu_base()
  par_O = par_Omega_beta()
  par_O$pSig$cc = 1
  pinit = c(tildecounts=10.5, Sa=1/20, sz=0.1, aa=1, bb=0.2) 
  #nlm(countsllik, pinit, alpha=alpha, counts=counts, q=q)
  out = optim(pinit, llik_counts, 
              alpha=alpha, counts=counts, 
              par_Fmu=par_Fmu, par_O=par_O, 
              q=q, bvm=bvm)$par 
  
  c(F_b(out[1], 13), F_Sa(out[2]), abs(out[3]), abs(out[4]), abs(out[5])) }
```




Finally,

```{r Fit it}
counts_MLE(alpha_list, counts_list) -> fits2
rbind(fits, fits1, fits2)
```